---
title: "Data exploration assignment"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

In this exercise, you will practice basic `dplyr` (the relevant package
within `tidyverse`) operations on a dataset of gene expression levels
across human tissues (from the GTEx Project). The dataset (`dt`) has one
row per gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

**Using HopGPT.** You are welcome to use HopGPT on this assignment. If
you do, add a couple of comment lines in that same code block noting
what you asked it and how you checked the answer. Remember that a
chatbot will happily give you code that runs perfectly and answers a
slightly different question than the one you asked.

For example:

```{r ai-comment-example, eval=FALSE}
# Asked HopGPT how to sort a tibble by a column with a space in its name.
# Checked: printed the first and last rows to confirm it sorted descending.
```

------------------------------------------------------------------------

### Start by loading the data into R:

Note the arguments, which specify that the first two lines of the file
will be skipped, that the file contains a header line that should be
used to form column names, and that the columns are tab (`\t`)
separated.

```{r load-data}
library(tidyverse)

dt <- read_delim("~/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

### Q1. How many genes are in the dataset, and how many tissues are represented?

Note that not every column is a tissue.

```{r q1}

# code and answer here
colnames(dt)
length(colnames(dt))
length(colnames(dt))-2

# answer = 54 tissues
```

------------------------------------------------------------------------

### Q2. How many genes are expressed above a given threshold in each tissue?

A common threshold is 1 TPM to define whether a gene is "expressed."\
Compute, for each of the three tissues, `Liver`, `Testis`, and
`Whole Blood`, how many genes are expressed above this threshold.\
Then identify which tissue has the **most** and which has the **fewest**
expressed genes. See this paper for one possible explanation: Xia et al.
2021: [PMC7891839](https://pmc.ncbi.nlm.nih.gov/articles/PMC7891839/).

Do the results change if you set the threshold higher, say at 10 TPM?

*Hint:* the column name `Whole Blood` contains a space, so it needs
backticks: `` dt$`Whole Blood` ``.

```{r q2}

# Code here
head(dt)

# Liver count
dt |> 
  select(Liver) |> 
  filter(Liver>10) |> 
  nrow()
  # 13499 for >1
  # 5132 for >10

# Testis count
dt |> 
  select(Testis) |> 
  filter(Testis>10) |> 
  nrow()
  # 25064 for >1 
  # 11471 for >10 

# Whole Blood Count
dt |> 
  select(`Whole Blood`) |> 
  filter(`Whole Blood`>10) |> 
  nrow()
  # 11616 for >1
  # 4482 for >10

# Interpretation here as a comment

#Liver = 13499 genes over TPM 1 and 5132 over TPM 10
#Testis = 25064 genes over TPM 1 and 11471 over TPM 10
#Whole Blood = 11616 genes over TPM 1 and 4482 over TPM 10

# Testis has the highest amount of genes expressed - the Whole Blood has the least amount of genes expressed


```

------------------------------------------------------------------------

### Q2b. Write a function

If you answered Q2 by copying the same line of code and editing the
tissue name, and then copying those lines again and editing the
threshold, you did six near-identical things by hand. Each time you
copied code represents a chance to introduce a mistake, and if you
wanted to change something, you would have to implement the change
multiple times.

Write a function `count_expressed()` that takes a vector of expression
values and a threshold, and **returns** the number of genes above that
threshold. Give the threshold a default value of 1.

Test it on a small vector where you already know the answer, such as
`c(0, 0.5, 1, 5, 100)`. Then use it to redo Q2 (all three tissues, both
thresholds) instead of writing out the counting code separately six
times.

```{r q2b}

# your function here

count_expressed <- function(exp_vals, thres=1) {
  filtered_vals <- exp_vals>thres
  #print(filtered_vals)
  num_of_genes <- sum(filtered_vals)
  return(num_of_genes)
}

# test on an input with a known answer
known_vector <- c(0, 0.5, 1, 5, 10)
count_expressed(known_vector)
count_expressed(known_vector,6)

# use it to answer Q2 again
count_expressed(dt$Liver, 1)
count_expressed(dt$Testis, 1)
count_expressed(dt$`Whole Blood`)
```

------------------------------------------------------------------------

### Q3. What are the 20 most highly expressed genes in the liver?

What do you notice about many of the genes in this list? Can you propose
a hypothesis to explain your observation?

```{r q3}

# Code here

top_liver <- dt |> 
  select(Description, Liver) |> 
  arrange(-Liver)

head(top_liver,20)
top_20_liver <- top_liver[1:20,]

# Interpretation here as a comment

## There is a large number of mitochondrial genes that have high expression in the liver.
#The liver is a major metabolism organ and mitochondrial function is upregulated in a metabolism functioning rich cells/region. I believe that the high metabolomic demand within liver tissue is the cause of the increased mito gene expression.

```

------------------------------------------------------------------------

### Q4. Compare the mean expression levels of mitochondrial versus nuclear-encoded genes in the Liver.

Genes encoded by the mitochondrial genome have the gene symbol prefix
`MT-`. The function `str_starts()` from the `stringr` package in
`tidyverse` tests whether a string (or a vector of strings) starts with
a given pattern, returning TRUE/FALSE (or a boolean vector) as output.

Use `str_starts()` to identify which genes are mitochondrially encoded
and store this information in a new Boolean column named
`is_mitochondrial`. Next, compute the mean and median expression of
mitochondrial genes in the liver, as well as the mean and median
expression of nuclear-encoded genes in the liver and compare them. Do
mitochondrial genes exhibit higher or lower mean and median expression
than nuclear genes? What does the difference between the mean and median
tell you about the distribution of gene expression?

**Check your pattern before you trust the result.** There is an easy
version of this that is wrong: matching `"MT"` instead of `"MT-"`. Count
how many genes each of the two patterns matches, and look at a few of
the genes that `"MT"` picks up but `"MT-"` does not. How much does the
mistake change your answer?

```{r q4}
# Code here

#mean of mito genes
mean_mito_expression <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-")) |> 
  filter(is.mitochondrial == TRUE) |> 
  summarize(mean(Liver))

#median of mito genes
median_mito_expression <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-")) |> 
  filter(is.mitochondrial == TRUE) |> 
  summarize(median(Liver))

#mean of nuc genes
mean_nuc_expression <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-")) |> 
  filter(is.mitochondrial != TRUE) |> 
  summarize(mean(Liver))

#median of nuc genes
median_nuc_expression <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-")) |> 
  filter(is.mitochondrial != TRUE) |> 
  summarize(median(Liver))

mean_mito_expression > mean_nuc_expression
median_mito_expression > median_nuc_expression
  
# how many genes match "MT-" vs "MT"? which genes differ?

MT_only_count <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-"),
         MT_only = str_starts(Description, "MT")
         ) |> 
  filter(MT_only == TRUE) |> 
  nrow()

MTdash_count <-top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-"),
         MT_only = str_starts(Description, "MT")
        ) |> 
  filter(is.mitochondrial == TRUE) |> 
  nrow()

MT_and_MTdash_count <- top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-"),
         MT_only = str_starts(Description, "MT")
        ) |> 
  filter(is.mitochondrial == TRUE & MT_only == TRUE) |> 
  nrow()

top_liver |> 
  mutate(is.mitochondrial = str_starts(Description, "MT-"),
         MT_only = str_starts(Description, "MT")
  ) |> 
  filter(MT_only == TRUE) |> 
  head(20)

#recalculate with MT only
#mean of MT genes
mean_MT_expression <- top_liver |> 
  mutate(MT_only = str_starts(Description, "MT")) |> 
  filter(MT_only == TRUE) |> 
  summarize(mean(Liver))

#median of MT genes
median_MT_expression <- top_liver |> 
  mutate(MT_only = str_starts(Description, "MT")) |> 
  filter(MT_only == TRUE) |> 
  summarize(median(Liver))

mean_MT_expression > mean_nuc_expression
median_MT_expression > mean_nuc_expression

# Interpretation here as a comment

  #Mito genes exhibit both a higher mean and median than nuclear genes in the liver. 

  #mean is the average and is more likely to be skewed by small groups of outliers, median is the exact middle data point in the set and is less likely to be skewed by outliers - more representative in large datasets.

  #MT- starting = 37
  #MT starting = 549

  # The mean of the MT starting gene expression is still higher but the median is lower than the nuclear genes

```

------------------------------------------------------------------------

### Q5. Which tissue is this?

So far we have told you which tissue to look at. Now you will find a
tissue based on the genes it expresses.

A gene is *tissue-specific* if it is expressed much more in one tissue
than it is on average across the body. To measure that, you first need
each gene's average expression across all tissues.

```{r mean-across-tissues}
# remove the columns that are not tissue expression
tissue_only <- dt %>%
  select(-Name, -Description)#, -is_mitochondrial)

# store the mean back in the original tibble
dt <- dt %>%
  mutate(mean_all_tissues = rowMeans(tissue_only))
```

**(a)** Pick any tissue in the dataset and compute a specificity score
for every gene:

$$\text{specificity} = \log_2 \left( \frac{\text{my tissue} + 1}{\text{mean across all tissues} + 1} \right)$$

Keep only genes expressed above 10 TPM in your tissue, sort by
specificity, and print the top 20 gene symbols.

**(b)** Give a classmate *only that list of 20 gene symbols* — not the
tissue name — and see whether they can identify the tissue. Do the same
for theirs. If the list doesn't give it away, explain why it
challenging.

**(c)** Why is the `> 10 TPM` filter in part (a)? Take it out and look
at what rises to the top. What kind of genes are the most \`\`specific"
when you do not require them to be meaningfully expressed?

**(d)** Housekeeping genes, which are highly expressed in every tissue,
should score near 0 on this specificity metric. Find a few genes with
specificity closest to 0 and high expression. Do they look like what you
would expect?

```{r mean-across tissues block #2}
  
#Bladder tissue

#specificity
dt <- dt |> 
  mutate(tissue_specificity_score = log2(((Bladder)+1)/((mean_all_tissues)+1)))

top_20_tissue_spec <- dt |> 
  filter(Bladder > 10) |> 
  arrange(-tissue_specificity_score) |> 
  head(20)

## a: Bladder tissue
## b: Super obvious via the sorting
## c: only 1 gene changed on the list. It was a urinary specific isoform of a gene already on the list
## d: Some were inactive small nuclear RNAs
```

------------------------------------------------------------------------

### Q6. Comparing gene expression between tissues

Another way to compare tissues is to ask how similar are their
expression patterns overall, across all genes?

R has a built-in function called `cor.test()` that measures the
correlation between two vectors of numbers. Remember that you can
extract the contents of a given column of a tibble as a vector using the
syntax `tibble_name$column_name`. If a column name contains spaces,
enclose the column name in backticks such as
`` tibble_name$`column name` ``.

Use `cor.test()` with the argument `method = "spearman"` to compute:

- the Spearman correlation between `Liver` and `Heart - Left Ventricle`
- the Spearman correlation between `Liver` and `Brain - Cortex`

Which pair of tissues is more strongly correlated? Does this result make
sense biologically?

You will see a warning about not being able to compute an exact p-value
because of ties. You can ignore it, as it does not affect the
qualitative results in this case. It is worth understanding where it
comes from, though: what is it about these data that creates so many
tied values?

```{r q6}
# Liver vs Heart answer
scorr_liv_heart <- cor.test(dt$Liver,dt$`Heart - Left Ventricle`, method = "spearman")
cor.test(dt$Liver,dt$`Heart - Left Ventricle`, method = "spearman")
# data:  dt$Liver and dt$`Heart - Left Ventricle`
#S = 3.0377e+12, p-value < 2.2e-16
#alternative hypothesis: true rho is not equal to 0
#sample estimates:
#  rho 
#0.8973187 

# Liver vs Brain answer
scorr_liv_bc <- cor.test(dt$Liver,dt$`Brain - Cortex`, method= "spearman")
cor.test(dt$Liver,dt$`Brain - Cortex`, method= "spearman")
#data:  dt$Liver and dt$`Brain - Cortex`
#S = 4.4259e+12, p-value < 2.2e-16
#alternative hypothesis: true rho is not equal to 0
#sample estimates:
#  rho 
#0.8503967 


# interpretation

# The liver is more correlated in gene expression to the heart than it is to the brain (if we are using the rho as the comparison value). The brain is an extremely specialized organ with incredibly specialized gene expression, visceral organs being more similar to each other than being similar to the central nervous system
```

------------------------------------------------------------------------

### Optional: further exploration

What other patterns are present in the data? Use tidyverse functions to
explore and summarize the data in different ways and see what you can
learn.

```{r optional}
# answer here
```
